---
title: "CiTonga Tone Production Analysis"
author:
  - David Nguyen
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 7,
  fig.height = 5
)
# Load required package
if (!requireNamespace("ggplot2", quietly=TRUE)) install.packages("ggplot2")
library(ggplot2)
````

# 1. Introduction

This analysis explores how the phonetic context of the first consonant (C1) affects the realization of High (H) and Low (L) tone F₀ values in Malawian CiTonga verb stems. We conduct two experiments:

* **Experiment A:** Compare mean F₀ of H-tone vowels following sonorant vs. voiced-obstruent onsets.
* **Experiment B:** Compare mean F₀ of H vs. L tones following each individual voiced-obstruent onset.

# 2. Data Loading

```{r load-data}
# Read the dataset
df <- read.csv("citonga.csv", stringsAsFactors = TRUE)
cat("Total tokens:", nrow(df), "\n")
str(df)
```

Columns of interest:

* `Tone` (“H” or “L”)
* `C1.class` (“sonorants” or “voiced obstruents”)
* `C1` (individual consonant: b, d, j, g, v, z, etc.)
* `meanf0` (mean pitch in Hz)

# 3. Experiment A: H Tone by C1 Class

## 3.1. Hypotheses

* **H₀:** Mean F₀ of H tones does not differ between sonorant and voiced-obstruent onsets.
* **H₁:** Mean F₀ of H tones differs between these onset classes.

## 3.2. Filter Data & Visualize

```{r expA-visualize}
df_H <- subset(df, Tone == "H" & C1.class %in% c("sonorants", "voiced obstruents"))
pA <- ggplot(df_H, aes(x = C1.class, y = meanf0)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(
    title = "Mean F₀ of H Tone by C1 Class",
    x     = "C1 Class",
    y     = "Mean F₀ (Hz)"
  ) +
  theme_minimal()
print(pA)
```

> Observation: The sonorant context appears to yield higher mean F₀ than the voiced-obstruent context.

## 3.3. Statistical Test

```{r expA-ttest}
sonorants_H  <- subset(df, Tone == "H" & C1.class == "sonorants")$meanf0
voicedObs_H  <- subset(df, Tone == "H" & C1.class == "voiced obstruents")$meanf0
testA <- t.test(sonorants_H, voicedObs_H, var.equal = TRUE)
testA
```

* **t** = `r round(testA$statistic, 4)`
* **df** = `r testA$parameter`
* **p-value** = `r signif(testA$p.value, 3)`

Since *p* < 0.05, we **reject** H₀ and conclude that onset class significantly affects H-tone F₀.

## 3.4. Conclusions from Experiment A

1. H-tone verbs preceded by sonorants have significantly higher pitch than those preceded by voiced obstruents.
2. Consonant class modulates tonal pitch realization in CiTonga.

# 4. Experiment B: Tone by Individual Voiced Obstruent

## 4.1. Filter Data & Visualize

```{r expB-visualize}
voiced_set <- c("b","d","j","g","v","z")
df_voiced <- subset(df, C1 %in% voiced_set)
pB <- ggplot(df_voiced, aes(x = C1, y = meanf0, fill = Tone)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Mean F₀ by Voiced Obstruent and Tone",
    x     = "Voiced Obstruent (C1)",
    y     = "Mean F₀ (Hz)"
  ) +
  theme_minimal()
print(pB)
```

> Observation: L tones often show higher mean F₀ than H tones after voiced obstruent onsets.

## 4.2. Statistical Test

```{r expB-ttest}
H_vo <- subset(df, Tone == "H" & C1 %in% voiced_set)$meanf0
L_vo <- subset(df, Tone == "L" & C1 %in% voiced_set)$meanf0
testB <- t.test(H_vo, L_vo, var.equal = TRUE)
testB
```

* **t** = `r round(testB$statistic, 4)`
* **df** = `r testB$parameter`
* **p-value** = `r signif(testB$p.value, 3)`

Since *p* < 0.05, we **reject** H₀ and conclude that tone category significantly affects F₀ in voiced-obstruent contexts.

## 4.3. Conclusions from Experiment B

1. After voiced obstruents, L-tone stems exhibit significantly higher pitch than H-tone stems.
2. This counterintuitive pattern suggests context-specific tonal enhancement or suppression mechanisms.

# 5. General Discussion & Follow-up

* Consonant context has a robust effect on tonal pitch realization in CiTonga.
* The reversal in Experiment B invites further phonetic and phonological investigation (e.g., aerodynamic influences, downstep patterns).
* **Follow-up question:** What articulatory or prosodic processes underlie the unexpected L-tone elevation?

# 6. References

* CiTonga field data (unpublished).
* Wickham H. (2016). *ggplot2: Elegant Graphics for Data Analysis*.
* R Core Team (2023). *R: A language and environment for statistical computing*.

# 7. Session Info

```{r session-info}
sessionInfo()
```
